A non-contextual hidden variable model for quantum mechanics

### 2016-07-12 finalPaper.tex
upadated the conclusions

\begin{document}
\title{A non-contextual hidden variable model for quantum mechanics}
\author{Atul Singh Arora}
\email{ms11003@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education and Research (IISER), Mohali}
\author{Arvind}
\email{arvind@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education
and Research (IISER), Mohali}
\begin{abstract}
We propose a non-contextual hidden variable model,
consistent with all predictions of quantum
mechanics (QM).  A careful scrutiny of
consistency requirements between any hidden
variable model and  quantum mechanics and the
corresponding no-go theorems, leads us to the
conclusion that the notion of contextuality is not
a necessary feature of QM.  The alternative view
that emerges, hinges on identifying a new classical
property, which we call ``multiplicativity''.  It turns
out that, by relaxing this condition of
``multiplicativity'' on hidden variable models, they
can be made  consistent with QM.  Advantages of
this view are illustrated by considering the
implications of non-multiplicativity to locality,
entanglement, superposition, and  apparent
contextuality. 
\end{abstract}
\pacs{}
\maketitle

Quantum mechanics has been one of the most
successful theories in physics so far, however,
there has not yet been a final word on its
completeness and interpretation~\cite{??}.
Einstein's~\cite{EinsteinEPR} work on the
incompleteness of QM and the subsequent seminal
work of Bell~\cite{Bell1964}, assessing the
compatibility of a more complete model involving
hidden variables (HV) and locality with QM, has
provided deep insights into  how the quantum world
differs from its classical counterpart.  In recent
times, these insights have been of pragmatic
utility in the areas of quantum information
processing, where EPR pairs are fundamental motifs
of entanglement~\cite{Ekert,PironioRndmnssCrtfcn}.
The work of  Kochen Specker~\cite{KochenSpecker}
and Gleason et.
al.~\cite{Gleason,BellOnHiddenVariables,Peres,Mermin})
broadened the schism between  HV models and QM.
They showed that it was contextuality and not
non-locality which was at the heart of this schism
and the incompatibility between HV models and QM
can arise even for a single quantum system.
Contextuality has thus been identified as a
fundamental non-classical feature of the quantum
world and experiments have also been proposed and
conducted to this effect~\cite{SimonContExpProp,
HuangContExp,YangContExp,HasegawaContExp}.
Contextuality, on the one hand has led to
investigations on the foundational aspects of
QM~\cite{PawelCntxClsscl,?????}, and on the other
hand has been harnessed for computation and
cryptography~\cite{HowardCntxCmptn,CabelloCntxScrt,??}.
Quantitative measures of contextuality based on
memory~\cite{MatthiasCntxMmry} have also been used
to demonstrate the completeness of
QM~\cite{CabelloMmryQM}.  However, not all HV
models are incompatible with QM, and a very
important case in point is  Bohm's model based on
precise trajectories for quantum
systems~\cite{Bohm1,Bohm2}.

@@

To bring out the non-multiplicativity of our model 
and to convince the reader of its generic validity,
we apply it to the Peres Mermin situation of two
spin-half particles. Consider the initial state of
the system 
$\left|\psi_{1}\right\rangle
=\left|00\right\rangle $. Assume we 
obtained $c=0.4$ as a random choice. To arrive
at the assignments, note that
$\left|00\right\rangle $ is an eigenket of only
$\hat{R}_{i},\hat{C}_{j}$ and
$\hat{A}_{33}=\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$.
Thus, in the first iteration, all these should be
assigned their respective eigenvalues. The
remaining operators must be assigned $-1$ as one
can readily verify by explicitly finding the
smallest $a$ as described in postulate 2 of the
model (see \tblref{HVmodel}). Two remarks are in
order. First, this model is manifestly
\emph{non-multiplicative}, for
$M_{1}(\hat{C}_{3})=1\neq
M_{1}(\hat{A}_{13})M_{1}(\hat{A}_{23})M_{1}(\hat{A}_{33})=-1$,
where the subscript has been introduced to index
the iteration. More precisely,
$M_{1}(\hat{O}):=M(\hat{O},\left[\left|\psi_{1}\right\rangle
=\left|00\right\rangle \right])$ where the
complete state $\left[\left|\psi_{1}\right\rangle
\right]$ implicitly refers to both the quantum
state $\left|00\right\rangle $ and the HV $c=0.4$.
Second, we must impose \emph{sequential
multiplicativity} as a consistency check of the
model, which in particular entails that
$M_{1}(\hat{C}_{3})=M_{1}(\hat{A}_{33})M_{2}(\hat{A}_{23})M_{3}(\hat{A}_{13})$,
where
$M_{2}:=M(\hat{O},\left[\left|\psi_{2}\right\rangle
\right])$,
$M_{3}:=M(\hat{O},\left[\left|\psi_{3}\right\rangle
\right])$ and $\left|\psi_{2}\right\rangle
,\,\left|\psi_{3}\right\rangle $ are obtained from
postulate 3. Note that for each iteration, a new
HV is generated. To illustrate \emph{sequential
multiplicativity}, we must choose to measure
$\hat{A}_{33}$. According to step 3, since
$\left|00\right\rangle $ is an eigenstate of
$\hat{A}_{33}$, the final state remains
$\left|00\right\rangle $.

For the next iteration, $i=2$, viz.
after the first measurement has been performed,
say the random number generator yielded $c=0.1$.
Since $\left|\psi\right\rangle $ is also unchanged
the assignment remains invariant (in fact any of
$c<0.5$ would yield the same result), which should
be evident from the previous exercise. For the
final step we choose to measure
$\hat p := \hat{A}_{23}(=\hat{\sigma}_{y}\otimes\hat{\sigma}_{y})$,
to proceed with sequentially measuring
$\hat{C}_{3}$. To simplify calculations, we note
\[ \left|00\right\rangle
=\frac{(\left|\tilde{+}\tilde{-}\right\rangle
+\left|\tilde{-}\tilde{+}\right\rangle
)/\sqrt{2}+(\left|\tilde{+}\tilde{+}\right\rangle
+\left|\tilde{-}\tilde{-}\right\rangle
)/\sqrt{2}}{\sqrt{2}}, \] where
$\left|\tilde{\pm}\right\rangle
=\left|0\right\rangle \pm i\left|1\right\rangle $
(eigenkets of $\hat{\sigma}_{y}$). Since
$\left|00\right\rangle $ is manifestly not an
eigenket of $\hat{p}$, we must find an appropriate
eigenket $\left|p_{-}\right\rangle $ s.t.
$\hat{p}\left|p_{-}\right\rangle
=-\left|p_{-}\right\rangle $, since $c=0.1$ and
$\left\langle p_{-}|00\right\rangle $ is already
$>0.1$. It is immediate that
$\left|p_{-}\right\rangle
=\left(\left|\tilde{+}\tilde{-}\right\rangle
+\left|\tilde{-}\tilde{+}\right\rangle
\right)/\sqrt{2}=\left(\left|00\right\rangle
+\left|11\right\rangle \right)/\sqrt{2}$, which
becomes the final state. \\ For the final
iteration, $i=3$, say we obtain $c=0.7$. So far,
we have $M_{1}(\hat{A}_{33})=1$ and
$M_{2}(\hat{A}_{23})=-1$. We must obtain
$M_{3}(\hat{A}_{13})=1$, independent of the value
of $c$, to be consistent. Let's check that.
Indeed, according to postulate 2, since
$\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}\left(\left|00\right\rangle
+\left|11\right\rangle
\right)/\sqrt{2}=1\left(\left|00\right\rangle
+\left|11\right\rangle \right)/\sqrt{2}$,
$M_{3}(\hat{A}_{13})=1$ for all allowed values of
$c$. As a remark, it maybe emphasised that the
$M_{2}(\hat{A}_{33})=M_{3}(\hat{A}_{33})$ and
$M_{2}(\hat{A}_{23})=M_{3}(\hat{A}_{23})$, which
essentially expresses compatibility of these
observables, viz. measurement of $\hat{A}_{13}$
doesn't affect the result one would obtain by
measuring operators compatible to it (granted they
have been measured once before).

%\begin{widetext}
%\onecolumngrid
\begin{table*}
 \begin{equation*}
\begin{array}{ccc} 
i=1,c=0.4,\left|\psi_{\text{init}}\right\rangle=\left|00\right\rangle & i=2,c=0.1,\left|\psi_{\text{init}}\right\rangle=\left|00\right\rangle & i=3,c=0.7,\left|\psi_{\text{init}}\right\rangle=\frac{\left|00\right\rangle +\left|11\right\rangle}{\sqrt{2}} \\
%\text{Iteration} & i=1 & i=2 & i=3\\ 
%\left|\psi_{\text{init}}\right\rangle  & \left|00\right\rangle  & \left|00\right\rangle  & \frac{\left|00\right\rangle +\left|11\right\rangle}{\sqrt{2}}\\ 
%\text{HV} & c=0.4 & c=0.1 & c=0.7\\ \\

@@

% \text{(Assignments)} &
% m_{1}(\hat{R}_{i}),m_{1}(\hat{C}_{j})=+1\,(j\neq3)
% &
% m_{2}(\hat{R}_{i}),m_{2}(\hat{C}_{j})=+1\,(j\neq3)
% &
% m_{3}(\hat{R}_{i}),m_{3}(\hat{C}_{j})=+1\,(j\neq3)\\
% & m_{1}(\hat{C}_{3})=-1 & m_{2}(\hat{C}_{3})=-1 &
% m_{3}(\hat{C}_{3})=-1
% \end{bmatrixT}
\label{eq:toyModel}
\end{array}
% \label{eq:toyModel}
\end{equation*}
\caption{HV model applied to the Peres Mermin situation}
\label{tbl:HVmodel}
\end{table*}
%\end{widetext}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
{\color{red} Will work on CC after everthin else
is done}
This has implications to Bohm's HV model as well, Bohmian Mechanics (BM), where the notion of positions (and hence trajectories) is precisely defined and particles are guided by the wavefunction. Consequently, given the initial conditions precisely (which is never the case experimentally) one can predict results of all experiments precisely. This entails that BM is capable of predicting outcomes of, for instance, a Stern Gerlach experiment. However, within BM two experiments which correspond to measurement of the same operator, can yield opposite results. Further, this can be demonstrated for a single spin half particle. This observation is often used as a demonstration of `contextuality' in BM, although we did not even speak of compatible measurements. It turns out, however, that if one restricts oneself to a one-one map between an experiment and an observable (by following some convention for example) then the so called `contextuality' can be removed from BM. One finds that the map so obtained from observables to their predictions turns out to be non-multiplicative, suggesting that at least in this case, non-multiplicativity is a more suitable explanation of non-classicality.

We have already constructed an explicit non-contextual model,
which is consistent with QM.  This model we knew
had to be non-multiplicative.  We will see how
non-multiplicativity gives rise to what one might
confuse to mean contextuality.  Imagine $\hat{B}_{1}=\hat{\sigma}_{z}\otimes\hat{\mathbb{I}}=\left|00\right\rangle
\left\langle 00\right|+\left|01\right\rangle
\left\langle 01\right|-\left[\left|10\right\rangle
\left\langle 10\right|+\left|11\right\rangle
\left\langle 11\right|\right],$ $
\hat{B}_{2}=\hat{\mathbb{I}}\otimes\hat{\sigma}_{z}=\left|10\right\rangle
\left\langle 10\right|+\left|11\right\rangle
\left\langle 11\right|-\left[\left|00\right\rangle
\left\langle 00\right|+\left|01\right\rangle
\left\langle 01\right|\right],$ while we define
$ \hat{C}=f(\{\hat{B}_{i}\})
=0.\left|00\right\rangle \left\langle
00\right|+1.\left|01\right\rangle \left\langle
01\right|+2.\left|10\right\rangle \left\langle
10\right|+3.\left|11\right\rangle \left\langle
11\right|. $
% \begin{align*} \hat{C}&=f(\{\hat{B}_{i}\}) \\
% &=0.\left|00\right\rangle \left\langle
% 00\right|+1.\left|01\right\rangle \left\langle
% 01\right|+2.\left|10\right\rangle \left\langle
% 10\right|+3.\left|11\right\rangle \left\langle
% 11\right|.  \end{align*}
 $\hat{C}$ maybe viewed as
a function of $\hat{B}_{1}$, $\hat{B}_{2}$ and
other operators $\hat{B}_{i}$ which are
constructed to obtain a maximally commuting set. A
measurement of $\hat{C}$, will collapse the state
into one of the states which are simultaneous
eignkets of $B_{1}$ and $B_{2}$. Consequently,
from the observed value of $\hat{C}$, one can
deduce the values of $\hat{B}_{1}$ and
$\hat{B}_{2}$. Now consider
$\sqrt{2}\left|\chi\right\rangle
=\left|10\right\rangle +\left|01\right\rangle $,
for which $M_{1}(\hat{B}_{1})=1$, and
$M_{1}(\hat{B}_{2})=1$, using the C-ingle model,
with $c<0.5$. However, $M_{1}(\hat{C})=1$, from
which one can deduce that $B_{1}$ was $+1$, while
$B_{2}$ was $-1$.  This property itself, one may
be tempted call contextuality, viz.  the value of
$B_{1}$ depends on whether it is measured alone or
with the remaining $\{B_{i}\}$. However, it must
be noted that $B_{1}$ has a well defined value,
and so does $\hat{C}$. Thus by our accepted
definition, there's no contextuality. It is just
that $M_{1}(\hat{C})\neq
\tilde f(M_{1}(\hat{B}_{1}),M_{1}(\hat{B}_{2}),\dots)$,
viz. the theory is non-multiplicative. Note that
after measuring $\hat{C}$ however,
$M_{2}(\hat{B}_{1})=+1$ and
$M_{2}(\hat{B}_{2})=-1$ (for any value of $c$)
consistent with those deduced by measuring
$\hat{C}$.

It is known that for simultaneous eigenstates of
$\hat{B}_{i}$, multiplicativity must hold.
Consequently, any violation of multiplicativity
must arise from states that are superpositions (of
the simultaneous eigenkets).  Note, however, that
entanglement is not necessary to show a violation
since, for example, the PM test is a state
independent test where a separable state can be
used to arrive at a contradiction. Entanglement 
can be used to show non-locality of non-multiplicative
prediction maps using Bell's theorem.
\bibliographystyle{apsrev4-1}
\bibliography{references}

\end{document}

>>>>

\begin{document}
\title{A non-contextual hidden variable model for quantum mechanics}
\author{Atul Singh Arora}
\email{ms11003@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education and Research (IISER), Mohali}
\author{Arvind}
\email{arvind@iisermohali.ac.in}
\selectlanguage{british}%
\affiliation{Indian Institute of Science Education
and Research (IISER), Mohali}
\begin{abstract}
We propose a non-contextual hidden variable model,
consistent with all predictions of quantum
mechanics (QM).  A careful scrutiny of consistency
requirements between any hidden variable model and
quantum mechanics and the corresponding no-go
theorems, leads us to the conclusion that the
notion of contextuality is not a necessary feature
of QM.  The alternative view that emerges, hinges
on identifying a new classical property, which we
call ``multiplicativity''.  It turns out that, by
relaxing this condition of ``multiplicativity'' on
hidden variable models, they can be made
consistent with QM.  Connection of
non-multiplicativity to Bohm Model and how it
apparent contextuality is also discussed.
\end{abstract}
\pacs{}
\maketitle
Quantum mechanics has been one of the most
successful theories in physics so far, however,
there has not yet been a final word on its
completeness and interpretation~\cite{??}.
Einstein's~\cite{EinsteinEPR} work on the
incompleteness of QM and the subsequent seminal
work of Bell~\cite{Bell1964}, assessing the
compatibility of a more complete model involving
hidden variables (HV) and locality with QM, has
provided deep insights into  how the quantum world
differs from its classical counterpart.  In recent
times, these insights have been of pragmatic
utility in the areas of quantum information
processing, where EPR pairs are fundamental motifs
of entanglement~\cite{Ekert,PironioRndmnssCrtfcn}.
The work of  Kochen Specker~\cite{KochenSpecker}
and Gleason et.
al.~\cite{Gleason,BellOnHiddenVariables,Peres,Mermin})
broadened the schism between  HV models and QM.
They showed that it was contextuality and not
non-locality which was at the heart of this schism
and the incompatibility between HV models and QM
can arise even for a single quantum system.
Contextuality has thus been identified as a
fundamental non-classical feature of the quantum
world and experiments have also been proposed and
conducted to this effect~\cite{SimonContExpProp,
HuangContExp,YangContExp,HasegawaContExp}.
Contextuality, on the one hand has led to
investigations on the foundational aspects of
QM~\cite{PawelCntxClsscl,?????}, and on the other
hand has been harnessed for computation and
cryptography~\cite{HowardCntxCmptn,CabelloCntxScrt,??}.
Quantitative measures of contextuality based on
memory~\cite{MatthiasCntxMmry} have also been used
to demonstrate the completeness of
QM~\cite{CabelloMmryQM}.  However, not all HV
models are incompatible with QM, and a very
important case in point is  Bohm's model based on
precise trajectories for quantum
systems~\cite{Bohm1,Bohm2}.

@@

To bring out the non-multiplicativity of our model 
and to convince the reader of its generic validity,
we apply it to the Peres Mermin situation of two
spin-half particles. Consider the initial state of
the system 
$\left|\psi_{1}\right\rangle
=\left|00\right\rangle $. Assume we 
obtained $c=0.4$ as a random choice. To arrive
at the assignments, note that
$\left|00\right\rangle $ is an eigenket of only
$\hat{R}_{i},\hat{C}_{j}$ and
$\hat{A}_{33}=\hat{\sigma}_{z}\otimes\hat{\sigma}_{z}$.
Thus, in the first iteration, all these should be
assigned their respective eigenvalues. The
remaining operators must be assigned $-1$ as one
can readily verify by explicitly finding the
smallest $a$ as described in postulate 2 of the
model (see \tblref{HVmodel}). 
%Two remarks are in
This model is manifestly
\emph{non-multiplicative}, for
$M_{1}(\hat{C}_{3})=1\neq
M_{1}(\hat{A}_{13})M_{1}(\hat{A}_{23})M_{1}(\hat{A}_{33})=-1$,
where the subscript refers to the iteration
number.
More precisely,
$M_{1}(\hat{O}):=M(\hat{O},\left[\left|\psi_{1}\right\rangle
=\left|00\right\rangle \right])$ where the
complete state $\left[\left|\psi_{1}\right\rangle
\right]$ implicitly refers to both the quantum
state $\left|00\right\rangle $ and the HV $c=0.4$.
%Second, we must impose \emph{sequential
%multiplicativity} as a consistency check of the
%model, which in particular entails that
The model however  obeys the \emph{sequential
multiplicativity} as required for consistency,
namely,
$M_{1}(\hat{C}_{3})=M_{1}(\hat{A}_{33})M_{2}(\hat{A}_{23})M_{3}(\hat{A}_{13})$,
where
$M_{2}:=M(\hat{O},\left[\left|\psi_{2}\right\rangle
\right])$,
$M_{3}:=M(\hat{O},\left[\left|\psi_{3}\right\rangle
\right])$ and $\left|\psi_{2}\right\rangle
,\,\left|\psi_{3}\right\rangle $ are obtained from
postulate 3. Note that for each iteration, a new
HV is generated. To illustrate \emph{sequential
multiplicativity}, we must choose to measure
$\hat{A}_{33}$. According to step 3, since
$\left|00\right\rangle $ is an eigenstate of
$\hat{A}_{33}$, the final state remains
$\left|00\right\rangle $.

For the next iteration, $i=2$, 
after the first measurement is over
say the random number generator yielded $c=0.1$.
Since $\left|\psi\right\rangle $ is also unchanged
the assignment remains invariant (in fact any of
$c<0.5$ would yield the same result as
evident from the previous exercise). For the
final step we choose to measure
$\hat{A}_{23}(=\hat{\sigma}_{y}\otimes\hat{\sigma}_{y})$,
to proceed with sequentially measuring
$\hat{C}_{3}$. To simplify calculations, we note
$\left|00\right\rangle
=\frac{1}{\sqrt 2} \left( (\left|\tilde{+}\tilde{-}\right\rangle
+\left|\tilde{-}\tilde{+}\right\rangle
)/\sqrt{2}+(\left|\tilde{+}\tilde{+}\right\rangle
+\left|\tilde{-}\tilde{-}\right\rangle
)/\sqrt{2}\right) , 
$
with
$\left|\tilde{\pm}\right\rangle
=\left|0\right\rangle \pm i\left|1\right\rangle $
(eigenkets of $\hat{\sigma}_{y}$). Since
$\left|00\right\rangle $ is manifestly not an
eigenket of $A_{23}$, we must find an appropriate
eigenket $\left|a^-_{23}\right\rangle $ s.t.
$A_{23}\left|a^{-}_{23}\right\rangle
=-\left|a^{-}_{23}\right\rangle $, since $c=0.1$ and
$\left\langle a_{23}^{-}|00\right\rangle $ is already
$>0.1$. It is immediate that
$\left|a^{-}_{23}\right\rangle
=\left(\left|\tilde{+}\tilde{-}\right\rangle
+\left|\tilde{-}\tilde{+}\right\rangle
\right)/\sqrt{2}=\left(\left|00\right\rangle
+\left|11\right\rangle \right)/\sqrt{2}$, which
becomes the final state.

 For the final
iteration, $i=3$, say we obtain $c=0.7$. So far,
we have $M_{1}(\hat{A}_{33})=1$ and
$M_{2}(\hat{A}_{23})=-1$. We must obtain
$M_{3}(\hat{A}_{13})=1$, independent of the value
of $c$, to be consistent. Let's check that.
Indeed, according to postulate 2, since
$\hat{\sigma}_{x}\otimes\hat{\sigma}_{x}\left(\left|00\right\rangle
+\left|11\right\rangle
\right)/\sqrt{2}=1\left(\left|00\right\rangle
+\left|11\right\rangle \right)/\sqrt{2}$,
$M_{3}(\hat{A}_{13})=1$ for all allowed values of
$c$. It is to be noted that
$M_{2}(\hat{A}_{33})=M_{3}(\hat{A}_{33})$ and
$M_{2}(\hat{A}_{23})=M_{3}(\hat{A}_{23})$, which
essentially expresses the compatibility of these
observables; i.e. 
once measured, the values of observables compatible
with $\hat{A}_{13}$ are not affected by the 
measurement of $\hat{A}_{13}$.

\begin{table*}
 \begin{equation*}
\begin{array}{ccc} 
i=1,c=0.4,\left|\psi_{\text{init}}\right\rangle=\left|00\right\rangle & i=2,c=0.1,\left|\psi_{\text{init}}\right\rangle=\left|00\right\rangle & i=3,c=0.7,\left|\psi_{\text{init}}\right\rangle=\frac{\left|00\right\rangle +\left|11\right\rangle}{\sqrt{2}} \\
%\text{Iteration} & i=1 & i=2 & i=3\\ 
%\left|\psi_{\text{init}}\right\rangle  & \left|00\right\rangle  & \left|00\right\rangle  & \frac{\left|00\right\rangle +\left|11\right\rangle}{\sqrt{2}}\\ 
%\text{HV} & c=0.4 & c=0.1 & c=0.7\\ \\

@@

% \text{(Assignments)} &
% m_{1}(\hat{R}_{i}),m_{1}(\hat{C}_{j})=+1\,(j\neq3)
% &
% m_{2}(\hat{R}_{i}),m_{2}(\hat{C}_{j})=+1\,(j\neq3)
% &
% m_{3}(\hat{R}_{i}),m_{3}(\hat{C}_{j})=+1\,(j\neq3)\\
% & m_{1}(\hat{C}_{3})=-1 & m_{2}(\hat{C}_{3})=-1 &
% m_{3}(\hat{C}_{3})=-1
% \end{bmatrixT}
\label{eq:toyModel}
\end{array}
% \label{eq:toyModel}
\end{equation*}
\caption{HV model applied to the Peres Mermin situation}
\label{tbl:HVmodel}
\end{table*}
This has implications to Bohm's HV model(BHVM),
where we have the notion of precise trajectories
for particle guided by the wavefunction.  In BHVM
if the initial conditions are precisely known the
entire trajectory of a particle can be predicted
including the individual outcome of measurements.
For the case of a single spin-1/2 particle in a
Stern Gerlach experiment, BHVM can predict
opposite results correspond to measurement of the
same operator.  This observation is often used as
a demonstration of `contextuality' in BM, although
we did not even speak of compatible
measurements\cite{??}.  It turns out, however,
that if one restricts oneself to a one-one map
between an experiment and an observable (by
following a certain convention), the so called
`contextuality' can be removed from BHVM.  One
finds that the map so obtained from observables to
their predictions turns out to be
non-multiplicative, suggesting that at least in
this case, non-multiplicativity is a more suitable
explanation of non-classicality.

We have  constructed an explicit non-contextual HV
model consistent with QM which is
non-multiplicative. We illustrate how
non-multiplicativity gives rise situations which
could get confused with presence of contextuality.

Imagine
$
\hat{B}_{1}=\hat{\sigma}_{z}\otimes\hat{\mathbb{I}}=
\left|00\right\rangle
\left\langle 00\right|+\left|01\right\rangle
\left\langle 01\right|-\left[\left|10\right\rangle
\left\langle 10\right|+\left|11\right\rangle
\left\langle 11\right|\right],
$ 
$
\hat{B}_{2}=\hat{\mathbb{I}}\otimes\hat{\sigma}_{z}=
\left|10\right\rangle
\left\langle 10\right|+\left|11\right\rangle
\left\langle 11\right|-\left[\left|00\right\rangle
\left\langle 00\right|+\left|01\right\rangle
\left\langle 01\right|\right],
$ 
while we define 
$
\hat{C}=f(\{\hat{B}_{i}\})
=0.\left|00\right\rangle \left\langle
00\right|+1.\left|01\right\rangle \left\langle
01\right|+2.\left|10\right\rangle \left\langle
10\right|+3.\left|11\right\rangle \left\langle
11\right|. 
$
$\hat{C}$ maybe viewed as a function of
$\hat{B}_{1}$, $\hat{B}_{2}$ and other operators
$\hat{B}_{i}$ which are constructed to obtain a
maximally commuting set. A measurement of
$\hat{C}$, will collapse the state into one of the
states which are simultaneous eignkets of $B_{1}$
and $B_{2}$. Consequently, from the observed value
of $\hat{C}$, one can deduce the values of
$\hat{B}_{1}$ and $\hat{B}_{2}$. Now consider
$\sqrt{2}\left|\chi\right\rangle
=\left|10\right\rangle +\left|01\right\rangle $,
for which $M_{1}(\hat{B}_{1})=1$, and
$M_{1}(\hat{B}_{2})=1$, using our model,
with $c<0.5$. However, $M_{1}(\hat{C})=1$, from
which one can deduce that $B_{1}$ was $+1$, while
$B_{2}$ was $-1$.  This property itself, one may
be tempted call contextuality, viz.  the value of
$B_{1}$ depends on whether it is measured alone or
with the remaining $\{B_{i}\}$. However, it must
be noted that $B_{1}$ has a well defined value,
and so does $\hat{C}$. Thus by our accepted
definition, there's no contextuality. It is just
that $M_{1}(\hat{C})\neq \tilde
f(M_{1}(\hat{B}_{1}),M_{1}(\hat{B}_{2}),\dots)$,
viz. the theory is non-multiplicative. Note that
after measuring $\hat{C}$ however,
$M_{2}(\hat{B}_{1})=+1$ and
$M_{2}(\hat{B}_{2})=-1$ (for any value of $c$)
consistent with those deduced by measuring
$\hat{C}$.

Evidently, Multiplicativity must hold for the
common eigenkets of $\hat{B}_{i}$.  Consequently,
any violation of multiplicativity must arise from
states that are superpositions or linear
combinations of these common eigenkets.  Note,
however, that entanglement is not necessary to
show a violation for instance the Peres Mermin
test is a state independent test where a separable
state can be used to arrive at a contradiction.
Entanglement can be used to show non-locality of
non-multiplicative prediction maps using Bell's
theorem  however, the notion non-multiplicativity
is more basic.  \begin{acknowledgments} Arvind
acknowledges funding from DST India under Grant
No. EMR/2014/000297.  \end{acknowledgments}
%\bibliographystyle{apsrev4-1}
\bibliography{references} \end{document}
